# Introductory Analysis of Storm Weather Effects on Population and Economy

## Synopsis
I look at historic data as made available by the US National Weather Service in order to assess the effects of weather phenomena on the general health of the population and also the economic impact. An exploratory analysis reveals that tornadoes and in general wind related phenomena are the most harmful both in terms of injuries and casualties and resulting as well in the highest financial loss.

## Data processing
The data processing stage is about getting hold of the data and doing any necessary transformations which allows us to do the type of analysis we are looking for.

```{r DataProcessing}
setInternet2(use = TRUE) # Knitr needs this to run download.file on SSL
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "StormData.csv.bz2")
storm <- read.csv(bzfile("StormData.csv.bz2"), stringsAsFactors = FALSE)
```

As a first step I looked at the Property and Crop Damage units. These are stored in the PROPDMGEXP and CROPDMGEXP respectively. The expected values where supposed to be blank, K (for '000s), M (for millions) and B for (billions). This is what I found, respectively:

```{r ExponentsAnalysis}
cat("Distribution of units for Property Damage Exponent")
table(storm$PROPDMGEXP)
cat("Distribution of units for Crop Damage Exponent")
table(storm$CROPDMGEXP)
```
In order to do the analysis I consolidated magnitude and exponent into a single figure and I also decided to NA all the ones not within the expected values

```{r ConsolidateDamage}
damageExponent <- function(exponent) {
  if(exponent == 'K')
    exp <- 3
  else if(exponent == 'M')
    exp <- 6
  else if(exponent == 'B')
    exp <- 9
  else
    exp <- 0
  return(exp)
}
storm$PROPDMG[which(!storm$PROPDMGEXP %in% c("","K","M","B"))] <- NA
storm$CROPDMG[which(!storm$CROPDMGEXP %in% c("","K","M","B"))] <- NA
validPROPDMG <- !is.na(storm$PROPDMG)
validCROPDMG <- !is.na(storm$CROPDMG)
propDmgExp <- vapply(storm$PROPDMGEXP[validPROPDMG], damageExponent, numeric(1))
cropDmgExp <- vapply(storm$CROPDMGEXP[validCROPDMG], damageExponent, numeric(1))
storm$PROPDMG[validPROPDMG] <- storm$PROPDMG[validPROPDMG] * (10 ** propDmgExp)
storm$CROPDMG[validCROPDMG] <- storm$CROPDMG[validCROPDMG] * (10 ** cropDmgExp)

```
After that I did some exploratory analysis on Event type. The lowest hanging fruit seems to come from consolidating everything to uppercase

```{r cleanup evtype}
storm$EVTYPE <- toupper(storm$EVTYPE)
```

## Results

### Harm to humans

For this preliminary study I decided to consider both Injuries and Fatalities as harm to the population. I present a table with the Top 10 events by harm to humans and a dot chart diagram for visualisation with harm as as defined by the previous sentence. Clearly Tornadoes are orders of magnitude more significant.

```{r results harm}
harm <- aggregate((INJURIES + FATALITIES) ~ EVTYPE, data = storm, sum)
colnames(harm) <- c("EVTYPE","AFFECTED")
harm <- harm[which(harm$AFFECTED > 0),]
harmTop10 <- head(harm[order(-harm$AFFECTED),], n = 10)
harmTop10
dotchart(harmTop10$AFFECTED, harmTop10$EVTYPE, main = "Top Phenomena by harm to humans")
```

### Economic impact

Things are not so clear cut, although wind related events seem to be the most financially destructive with Tornadoes against in the lead outstanding position.

```{r results economic}
loss <- aggregate((PROPDMG + CROPDMG) ~ EVTYPE,
                  data = storm, sum, na.rm = TRUE)
colnames(loss) <- c("EVTYPE", "ECOLOSS")
loss <- loss[which(loss$ECOLOSS > 0),]
lossTop10 <- head(loss[order(-loss$ECOLOSS),], n = 10)
lossTop10
dotchart(lossTop10$ECOLOSS, lossTop10$EVTYPE, main = "Top Phenomena by economic loss")
```