---
title: "Introductory Analysis of Storm Weather Effects on Population and Economy"
author: "Arturo Pina"
date: "Monday, July 20, 2015"
output: html_document
---
# Introductory Analysis of Storm Weather Effects on Population and Economy

## Synopsis

## Data processing
The data processing stage is about getting hold of the data and doing any necessary transformations which allows us to do the type of analysis we are looking for.

```{r DataProcessing}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "StormData.csv.bz2")
storm <- read.csv(bzfile("StormData.csv.bz2"), stringsAsFactors = FALSE)
```

As a first step I looked at the Property and Crop Damage units. These are stored in the PROPDMGEXP and CROPDMGEXP respectively. The expected values where supposed to be blank, K (for '000s), M (for millions) and B for (billions). This is what I found, respectively:

```{r ExponentsAnalysis}
cat("Distribution of units for Property Damage Exponent")
table(storm$PROPDMGEXP)
cat("Distribution of units for Crop Damage Exponent")
table(storm$CROPDMGEXP)
```
In order to do the analysis I consolidated magnitude and exponent into a single figure and I also decided to NA all the ones not within the expected values

```{r ConsolidateDamage}
damageExponent <- function(exponent) {
  if(exponent == 'K')
    exp <- 3
  else if(exponent == 'M')
    exp <- 6
  else if(exponent == 'B')
    exp <- 9
  else
    exp <- 0
  return(exp)
}
storm$PROPDMG[which(!storm$PROPDMGEXP %in% c("","K","M","B"))] <- NA
storm$CROPDMG[which(!storm$CROPDMGEXP %in% c("","K","M","B"))] <- NA
validPROPDMG <- !is.na(storm$PROPDMG)
validCROPDMG <- !is.na(storm$CROPDMG)
propDmgExp <- vapply(storm$PROPDMGEXP[validPROPDMG], damageExponent, numeric(1))
cropDmgExp <- vapply(storm$CROPDMGEXP[validCROPDMG], damageExponent, numeric(1))
storm$PROPDMG[validPROPDMG] <- storm$PROPDMG[validPROPDMG] * (10 ** propDmgExp)
storm$CROPDMG[validCROPDMG] <- storm$CROPDMG[validCROPDMG] * (10 ** cropDmgExp)

```
After that I did some exploratory analysis on Event type. The lowest hanging fruit seems to come from consolidating everything to uppercase

```{r cleanup evtype}
storm$EVTYPE <- toupper(storm$EVTYPE)
```

## Results

```{r results}
harm <- aggregate((INJURIES + FATALITIES) ~ EVTYPE, data = storm, sum)
colnames(harm) <- c("EVTYPE","AFFECTED")
harm <- harm[which(harm$AFFECTED > 0),]
harm[order(-harm$AFFECTED),]
harmTop10 <- head(harm[order(-harm$AFFECTED),], n = 10)
dotchart(harmTop10$AFFECTED, harmTop10$EVTYPE)
```
